Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns

### 2019-08-10 ms/manuscript.Rmd
Cut text from statistical analysis section

We calculated daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each site from the raw climate data, then overall means from the daily values. We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (hereafter, "climatic variables"). Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models. We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate. We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone and their interactions, then selecting the one with the lowest corrected Akaike Information Criterion (AIC~c~) [@Akaike1973]. A single outlier site with unusually high temperature and low humidity (Mt. Rotui, 830 m) site was excluded from ANCOVA and linear models. Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island (J. Nitta, pers. obs.).

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea (*n* = `r length(species_list)` \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ("strict" dataset).

@@

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using methods that can test for significant differences between traits associated with a binary character while accounting for phylogeny. For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait. For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012]. This method compares the likelihood between two alternative models for a pair of binary traits. The first model tests the null hypothesis that rates of evolution of the two traits are independent. The second model tests the alternative hypothesis that the rates of evolution of one trait depends on the other. The likelihoods of obtaining the observed data are calculated for each model, then compared using a log likelihood test. A significantly better fit of the dependent model indicates that evolution of the two traits is correlated. The same analyses were also done on the "strict" dataset.

@@

We used distance-based methods to analyze phylogenetic and functional diversity. Mean phylogenetic distance (MPD) and mean nearest taxon distance (MNTD) describe the mean distance between all species in a community (MPD) or the mean distance between the closest pairs of species in a community (MNTD) [@Webb2000]. These metrics were originally formulated for phylogenetic distances (thus "phylogenetic" in the name), but can be applied in the same way to trait-based distances, enabling a direct comparison between the two [@Sessa2018]. For functional diversity, we used Gower's dissimilarity coeffecient [@Gower1971] because it can use continuous, binary, and qualitative traits, and accepts missing data. We log-transformed traits as needed to account for skew, scaled each trait by its standard deviation, and excluded frond width and length (both frond size traits highly correlated with stipe length) before calculating the distance matrix using the \function{gowdis} function in the R pacakge \package{FD} [@Laliberte2011], with equal weighting for sporophyte and gametophyte traits. To test for the effect of different trait selection schemes, we compared the results of running the same analyses for various sets of traits with and without log-transformation (Supporting Information).

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities (equivalent to the Net Relatedness Index and Nearest Taxon Index, respectively, of @Webb2000 with the sign reversed). Selection of the randomization procedure can strongly bias outcomes [@Gotelli2006;@Ulrich2012]. Our null hypothesis is that fern spores are easily capable of dispersal between terrestrial and epiphytic habitats, and across the relatively short distances separating our plots. We therefore constructed null communities using the "independent swap" algorithm of @Gotelli2000 that maintains species frequency and abundance while randomizing species identities. We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available (`r Ntip(phy)` \si{\species} total; [@Nitta2017]) for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities. We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics. Hereafter, "MPD~phy~" and "MNTD~phy~", and "MPD~func~" and "MNTD~func~", will be used to refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

We tested for differences in diversity metrics between epiphytic and terrestrial communities with a paired \tval-test. We simultaneously tested the effects of climate and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data. Our dataset included `r nrow(div_metrics_all) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all climatic variables are related to elevation (see Results), we also constructed linear models for each diversity metric using the same model selection procedure as in the climate analysis. We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].

>>>>

We calculated daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each site, then overall means from the daily values. We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (hereafter, "climatic variables"). Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models. We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate. We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone and their interactions, then selecting the one with the lowest corrected Akaike Information Criterion (AIC~c~) [@Akaike1973]. A single outlier site with unusually high temperature and low humidity (Mt. Rotui, 830 m) site was excluded from ANCOVA and linear models. Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established (J. Nitta, pers. obs.).

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea (*n* = `r length(species_list)` \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under BM: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010]. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ("strict" dataset).

@@

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using methods that can test for significant differences between traits associated with a binary character while accounting for phylogeny. For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait. For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012]. This method compares the likelihood between two alternative models for a pair of binary traits. In the first model, rates of evolution of the two traits are independent. In the second model, the rate of evolution of one trait depends on the other. A significantly better fit of the second model indicates that evolution of the two traits is correlated. The same analyses were also done on the "strict" dataset.

@@

We used distance-based methods to analyze community phylogenetic and functional diversity. Mean phylogenetic distance (MPD) and mean nearest taxon distance (MNTD) describe the mean distance between all species in a community (MPD) or the mean distance between the closest pairs of species in a community (MNTD) [@Webb2000]. Although these metrics were originally formulated for phylogenetic distances, they can be applied in the same way to trait-based distances, enabling a direct comparison between the two [@Sessa2018]. We used Gower's dissimilarity coeffecient [@Gower1971] to calculate functional distances because it can use continuous, binary, and qualitative traits, and accepts missing data. We log-transformed traits as needed to account for skew, scaled each trait by its standard deviation, and excluded frond width and length (both frond size traits highly correlated with stipe length) before calculating the distance matrix using the \function{gowdis} function in the R pacakge \package{FD} [@Laliberte2011], with equal weighting for sporophyte and gametophyte traits. To test for the effect of different trait selection schemes, we compared the results of running the same analyses for various sets of traits with and without log-transformation (Supporting Information).

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities (equivalent to the Net Relatedness Index and Nearest Taxon Index, respectively, of @Webb2000 with the sign reversed). We constructed null communities using the "independent swap" algorithm of @Gotelli2000 that maintains species frequency and abundance while randomizing species identities. We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available (`r Ntip(phy)` \si{\species} total; [@Nitta2017]) for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities. We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics. Hereafter, "MPD~phy~" and "MNTD~phy~", and "MPD~func~" and "MNTD~func~", refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

We simultaneously tested the effects of climate and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data. Our dataset included `r nrow(div_metrics_all) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all climatic variables are related to elevation (see Results), we also constructed linear models for each diversity metric using the same model selection procedure as in the climate analysis. We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].
